---
title: "Cleaning"
---

## Goals of this notebook

The steps we'll take to prepare our data:

- Download the data
- Import it into our notebook
- Clean up data types and columns
- Export the data for next notebook

# Setup 

Loading the libraries 

```{r}
#| label: setup
#| message: false


library(tidyverse)
library(janitor)
```

## Downloading the Data 

In this section of the cleaning notebook, we're loading the VERSA program data. This data accounts for arrests made under Operation Lone Star starting in January 2023. 

Here is the summary of the differences between SPURS system and VERSA from the Depart of Public Safety Communications: 

"Please note that arrests are now documented on two spreadsheets due to the implementation of a new record management system in 2023 by the Texas Highway Patrol (THP). Prior to January 2023, THP utilized the Enterprise Management (EM) System for traffic enforcement activity (written warnings, citations, and CMV inspections) and SPURS for documenting entrusted property and criminal case reports. Beginning in January of 2023, THP underwent a statewide rollout of the Versaterm Software Suite, completed in August 2023, which serves as THP’s only RMS system for regular traffic enforcement activity, entrusted property, as well as general and arrest offense report entry." 

More information about Operation Lone Star is found in the index page of this website. 

The data was obtained by a Public Records Request from KUT. 

```{r}
download.file(
  "https://docs.google.com/spreadsheets/d/e/2PACX-1vT86k4riODyqAQeAQB0kB7NJlXDBQSJu19cUzNdJNZzUUXFLWuQHl5rmWZfXicGl2zqzztB6Hx8Gl-x/pub?output=csv","data-raw/versa_analysis.csv", mode = "wb"
)
```

## Importing the Data 

Here we're importing the VERSA data spreadsheet that I already have downloaded to my "data-raw" folder. Then, we're adding our data to a code chunk. For simplicity it will be called "versa_analysis," which we will then peak at in order to look at what our data looks like in order to clean. 


```{r}
versa_analysis<- read_csv("data-raw/versa_analysis.csv")

versa_analysis
```

```{r}
versa_analysis |> glimpse()
```


