---
title: "Cleaning"
---

## Goals of this notebook

The steps we'll take to prepare our data:

- Download the data
- Import it into our notebook
- Clean up data types and columns
- Export the data for next notebook

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
```

## Importing data

We're importing the data which was collected in a PIR by KUT's Lauren McGaughy. 

```{r}
spurs_raw <- read_csv("data-raw/spurs_raw_data.csv") |> clean_names()

spurs_raw
```


## Changing name

```{r}
spurs_renamed <- spurs_raw |> 
  rename(arrest_date_old = arrest_date,
         charge = offense,
         severity_code = level_group,
         arrest_officer = arrested_by,
         gender_code = person_sex,
         statute = citation,
         ethnicity_race = person_race
         )

spurs_renamed
```


## Changing dates

```{r}
spurs_date <- spurs_renamed |>
  mutate(
    charge_date = mdy_hm(arrest_date_old) |> date()
  )
```

## Changing officer column

```{r}
spurs_officer_cleaning <- spurs_date |> 
  mutate(new_officer = str_split_i(arrest_officer, " - ", 1), .after = arrest_officer,
         id_officer = str_split_i(arrest_officer, " - ", 2)
         )

spurs_officer_cleaning
```


## Changing ethnicity column

```{r}
spurs_race <- spurs_officer_cleaning |> 
  mutate(
    person_race_abbr = case_match(ethnicity_race, "White" ~ "W", "Hispanic" ~ "H", "Hispanic or Latino" ~ "H", "Black" ~ "B", "Asian" ~ "A"),
    person_gender_abbr = case_match(gender_code, "M - Male" ~ "M", "F - Female" ~ "F")
    )

spurs_race
```


## Changing names to uppercase

```{r}
spurs_uppercase <- spurs_race |> 
  mutate(person_first_name = str_to_upper(person_first_name),
         person_last_name = str_to_upper(person_last_name))

spurs_uppercase
```


## Creating a charge count
 
```{r}
spurs_charge_count <- spurs_uppercase |>
  group_by(arrest_date_old, arrest_county, arrest_division, arrest_id, arrest_state, arrest_urn, arrest_officer, new_officer, id_officer, statute, code, inv_number, severity_code, charge, person_first_name, person_last_name, ethnicity_race, gender_code, person_age, charge_date, person_race_abbr, person_gender_abbr ) |> 
  summarise(charge_count = n()) |> mutate(charge_count = as.numeric(charge))

spurs_charge_count
```


## Severity code

```{r}
spurs_code_clean <- spurs_charge_count |> 
  mutate(severity_code_clean = str_remove(severity_code, "SPURS "), .after = severity_code)

spurs_code_clean
```


## Selecting which columns I actually need

```{r}
spurs_clean <- spurs_code_clean |> ungroup() |> 
  select(
    arrest_county,
    arrest_division,
    arrest_id,
    arrest_state,
    arrest_urn,
    arresting_officer = new_officer,
    spurs_officer_id = id_officer,
    statute,
    code,
    inv_number,
    severity_code_clean,
    charge,
    person_first_name,
    person_last_name,
    person_age,
    charge_date,
    person_race_abbr,
    person_gender_abbr,
    charge_count
  )

spurs_clean |> glimpse()
```

## Exporting to data_processed

```{r}
spurs_clean |>
write_rds("data-processed/01-clean-spurs.rds")
```


